---
title: "Discovering Labor Patterns in Education"
---
Author: Jaimie Yu

Education plays a vital role in defining the experience of unemployment among worker populations, especially within disrupted economies. The four graphs included within this section combine to demonstrate that educational levels impacted not only the chances of job loss due to COVID-19 but also created challenges associated with reverting to that employment. By illustrating these points within various graphs, it becomes clear that educational factors were among the most prominent and stable factors within these labor market dynamics.

The first plot lays the groundwork with a comparison of unemployment duration based on levels of education. A very telling side-by-side comparison reveals that individuals with lower levels of education, including dropouts and high school graduates, had significantly longer unemployment durations compared with individuals who have college degrees. It thus lays down a groundwork that there existed a very strong relationship between levels of education and unemployment duration.

The second plot continues to develop this storyline by incorporating a temporal aspect. Rather than comparing various groups at a single point, it traces various levels of education from 2019 through 2024. As seen, while all groups were impacted simultaneously at the onset of COVID, their rate of recovery varied considerably. That is, highly educated workers had reached pre-COVID levels approximately within 2022, while those with lower levels of education experienced an irregular rate of recovery. Again, this specific plot addresses the question above and clearly shows that highly educated workers were the first to recover.

Third, there is a plot that focuses on the intersection of education with yet another significant demographic variable: veteran status. Based on unemployment duration for both veterans and nonveterans stratified by educational attainment, it clearly shows that education does not exist or interact solely on its own. It would appear that some people, like lower-educated veterans, have it doubly hard, while higher education reduces discrepancies between veterans and nonveterans.
## #1 Unemployment Duration by Education Level

This plot compares unemployment duration across workers with different levels of educational attainment. The boxplots display the distribution of unemployment spells for each education group, allowing us to observe both the spread and central tendencies. We also added a red dashed line indicating the overall median unemployment duration across all groups, which helps visually compare how each education category deviates from the national median.

Contrary to expectations that individuals with lower education would face the longest unemployment spells, the graph shows that high school graduates, individuals with some high school education but no diploma, and even those with a bachelor’s degree have median unemployment durations that fall above the overall median of roughly 10 weeks. This suggests that unemployment challenges during the COVID period did not fall strictly along a simple “more education equals shorter unemployment” pattern.

Why it matters: This visualization provides an important baseline for understanding how education level shapes labor-market vulnerability. While prior research and intuition often suggest that lower-educated workers experience longer unemployment spells, this plot shows that several groups across both lower and higher education categories were significantly affected.

This nuance is especially meaningful in the context of COVID-19 recovery. The pandemic disrupted industries unevenly, and this pattern of unemployment duration suggests that both lower-skilled and some higher-skilled workers faced prolonged difficulties reentering the workforce. Identifying these disparities helps policymakers and researchers better understand which educational groups were hit hardest and who may need targeted interventions to support employment recovery.

```{r setup, message=FALSE, warning=FALSE}
#| label: make-unemp-duration-edu
#| include: false
#| message: false
#| warning: false

source(here::here("src", "figs", "unemployment_duration_edulevel.R"))

```

```{r}
#| label: fig-unemp-duration-edu
#| fig-cap: "Unemployment duration by education level"
#| fig-align: "center"
#| out-width: "100%"
#| echo: false

knitr::include_graphics(
  here::here("figs", "unemp_by_edu.png")
)
```

## #2 Unemployment Duration Over Time by Education

This time-series plot tracks how unemployment duration changed for each education group from 2019 to 2024. The curves show how all groups experienced a sharp rise in unemployment duration in 2020, followed by very different recovery paths. College-educated workers recovered relatively quickly, while workers with only a high school degree or less saw a slower and more uneven decline. Interestingly, we saw another sharp increase for the group with doctorate degree. 

Why it matters: This plot shifts the analysis from levels to trajectories. It reveals which education levels “bounced back” the fastest and shows that the long-term effects of the pandemic were far more persistent for lower-educated workers.

```{r setup, message=FALSE, warning=FALSE}
#| label: make-unemp-duration-edu-overtime
#| include: false
#| message: false
#| warning: false
source(here::here("src", "figs", "unemployment_edu_time_plot.R"))
```

```{r}
#| label: fig-unemp-duration-edu-overtime
#| fig-cap: "Unemployment duration by education level overtime"
#| fig-align: "center"
#| out-width: "80%"
#| echo: false

knitr::include_graphics(
  here::here("figs", "unemp_edu_time_plot.png")
)
```

## #3 Unemployment Duration by Education and Veteran Status

Description: This plot compares unemployment duration for veterans and non-veterans within each education category. The median number for each group is written in each boxed plots. By showing how these two identities intersect, it uncovers differences that are not visible in education-only comparisons. For some groups—especially those with lower education—veterans may face longer unemployment spells than non-veterans. Higher education seems to narrow these gaps. Interestingly, we found that the veteran group with bachelor's degree has a harder time bouncing back form unemployment with a median of 12 weeks compared to the non-veteran group with bachelor's degree with a median of 8 weeks.

Why it matters: This plot shows that education interacts with veteran status in important ways. It demonstrates that educational disadvantage can be compounded by other demographic factors, revealing layers of inequality that influence unemployment duration.

```{r setup, message=FALSE, warning=FALSE}
#| label: unemployment-race-veteran status
#| include: false
#| message: false
#| warning: false
source(here::here("src", "figs", "edu_vet_plot.R"))
```

```{r}
#| label: unemploymentduration_race_vetstatus
#| fig-cap: "unemployment duration by education and veteran status"
#| fig-align: "center"
#| out-width: "80%"
#| echo: false

knitr::include_graphics(
  here::here("figs", "unemp_edu_veteran.png")
)
```
