---
---

## **Discovering Patterns in Veteran Status**

One of our key demographic variables is a person's veteran status. It determines our study's scope and impact for our target population and audience of interest, the United States Government. The following visualizations establish key information on the population of survey respondents that are veterans. 

```{r setup, message=FALSE, warning=FALSE}
#| label: make-count-vet
#| include: false
#| message: false
#| warning: false

source(here::here("src", "figs", "final_plotting_script_daniella.R"))

```

```{r}
#| label: fig-countvet
#| fig-cap: "Figure 1: Count of Respondents and their Veteran status."
#| fig-align: "center"
#| out-width: "100%"
#| echo: false

knitr::include_graphics(
  here::here("src", "figs", "Total_Respondents_Across_Years.png")
)
```

In 2023, the United States reported having a total of 15.8 American veterans. They receive support from the United States Department of Government Affairs, including services like housing, healthcare, and employment support. Given it is a large federal department, they budgeted 441.3 billion for the 2026 Fiscal Year. 

Figure 1 demonstrates the amount of respondents and their status as either a veteran or having never served for the United States government. The Community Population Survey (CPS) fails to capture all of the household who identify as veterans or non-veterans. They take a sample of American households, meaning that the number of veterans reported by CPS is much less than the government's reported number. CPS respondents are largely non-veterans or of unknown status. Notably, the number of respondents in 2023 increased for both groups, while veteran respondents remained largely stable.

```{r setup, message=FALSE, warning=FALSE}
#| label: make-unemp-duration-race
#| include: false
#| message: false
#| warning: false

source(here::here("src", "figs", "final_plotting_script_daniella.R"))

```

```{r}
#| label: fig-unemp-duration-race
#| fig-cap: "Figure 2: Count of Respondents and their Veteran and Race Demographics"
#| fig-align: "center"
#| out-width: "100%"
#| echo: false

knitr::include_graphics(
  here::here("src", "figs", "Total_Respondants_Across_Year_Divded_by_Race_and_Veteran_Status.png")
)
```

Figure 2 gives us more information about the two groups of interest, non-veterans and veterans. The visualization further divides the variables by the race of the respondents, breaking it down into white and non-white. This study will discuss the role of race further in a separate page. Figure 2 emphasizes the discrepancies among the different demographics in the CPS respondents. White groups are the largest group responding to CPS and Non-White respondents are the fewest. 

```{r setup, message=FALSE, warning=FALSE}
#| label: make-uemp-duration-density
#| include: false
#| message: false
#| warning: false
source(here::here("src", "figs", "final_plotting_script_daniella.R"))
```

```{r}
#| label: fig-unemploymentduration-race-vetstatus
#| fig-cap: "Figure 3: Unemployment Duration (in weeks) Density Distribution Across Veteran Status"
#| fig-align: "center"
#| out-width: "70%"
#| echo: false

knitr::include_graphics(
  here::here("src","figs", "Unemployment_Duration Distribution_by_Veteran_Status.png")
)
```

Returning to the analysis of the veteran and non-veteran variable, Figure 3 compares the two groups and the shape of the data across their unemployment duration. Both groups have a larger density of respondents report their unemployment lasting under 10 weeks. From there, the density decreases until a slight increase at 50 weeks. The non-veteran group has slightly more variation after the 10 week mark. 

```{r setup, message=FALSE, warning=FALSE}
#| label: unemployment-nine-weeks
#| include: false
#| message: false
#| warning: false
source(here::here("src", "figs", "final_plotting_script_daniella.R"))
```

```{r}
#| label: fig-unemploymentduration-pro
#| fig-cap: "Figure 4: Proportion of People Unemployed for 9 Weeks, 2020-2024"
#| fig-align: "center"
#| out-width: "70%"
#| echo: false

knitr::include_graphics(
  here::here("src","figs", "Median_proportion_plot.png")
)
```

Following figure 3, the team discovered that the median weeks of unemployment duration, across both groups, is 9 weeks. Figure 4 works off this finding to see how the proportion of people who are either veterans or not changes across our time frame of interest at the 9 week mark. The majority of people unemployed for 9 weeks are non-veterans, this is to be expected as the majority of respondents are non-veterans. Interestingly, the proportion of veterans unemployed for 9 weeks noticeably increases in 2022, with slightly under one fourth of respondents being veterans.

The remaining analysis will dive into more depth about how unemployment impacts different demographic groups and connecting it back to the variable of veteran status. 

\
